Ticketing and Turnout: The Participatory Consequences of Low-Level Police Contact

### 2022-04-19 markdown/draft_paper.Rmd
submissions

---
title: "Ticketing and Turnout: How Policing for Profit Shapes Democratic Participation"

author:
- Jonathan Ben-Menachem^[PhD Student, Columbia University, Department of Sociology (jb4487@columbia.edu)]
- Kevin Morris^[PhD Student, CUNY Graduate Center, Department of Sociology (kmorris@gradcenter.cuny.edu)]
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    keep_tex: true
bibliography: "policing_turnout.bib"
link-citations: yes
fontsize: 12pt
header-includes:
    - \usepackage{rotating}
    - \usepackage{setspace}
abstract: |
  The American criminal legal system is an important site of political socialization: sociologists show how the alienating effect of criminalization pushes people away from public institutions, and political scientists have measured the effects of policing and incarceration on political participation. Despite this burgeoning literature, no research has directly investigated how police ticketing practices affect political participation. We use three approaches to answer this question: we leverage individual-level administrative ticketing data from Hillsborough County, Florida between 2010 and 2020, a city-level analysis using Census of Governments data and the national voter file, and national survey data. We use a genetic matching algorithm and difference-in-differences to estimate that being ticketed decreases an individual’s likelihood of voting in federal elections by 1.7 percentage points, but depressive effects are smaller for Black residents. Nationally, we estimate that each 10 percent increase in city fines and fees collections between 2012 and 2017 was associated with a roughly 0.04 percentage point turnout decrease but was not associated with any change in Black turnout. The survey data show that Black people who report that they or a family member were recently stopped by police are more likely to vote than those who reported no such contact.

  |
  |
  **Keywords**: voting, monetary sanctions, race, punishment, political participation
---

@@

Work exploring how criminalization directly and indirectly influences political participation has exploded in recent years following path-breaking work from political sociologists Chris Uggen and Jeff Manza [@Uggen2002; @Manza2006]. Scholars have found that criminal legal contact (i.e. arrest, conviction, incarceration) consistently discourages voting [@Weaver2010; @Burch2011; @White2019a]. Such work has largely focused on the effects of highly disruptive contact with the criminal legal system such as incarceration and felony convictions [@Burch2014; @Lee2014]. While ticketing involves potentially negative interactions with the state, it does not necessarily carry the disruptive consequences of a felony conviction and might thus politicize Americans in unique ways. Building on recent theoretical work from @Walker2020a and an analysis of the historical context of fees and fines collections in the United States, this paper is the first to explicitly theorize how local ticketing practices shape the political socialization of ticketed individuals. We argue that ticketing depresses voter turnout overall, but also catalyzes a “linked fate” political analysis among Black voters that either causes increased turnout or mitigates the turnout reductions observed among non-Black individuals.

Additionally, our project represents the first use of individual-level administrative data to identify the causal effect of ticketing on voter behavior. The use of administrative data marks an important step forward in our understanding of how low-level contact with the criminal legal system structures political participation. Past work looking at the individual-level effects of low-level contact has relied entirely on survey or qualitative data [e.g. @Walker2014; @Weaver2010]. Existing research allows for the testing of specific psychological mechanisms and personal interpretations of criminal legal contact, but does not allow us to generalize more broadly. As Weaver and Lerman [-@Weaver2010: 821] note, it may also introduce measurement bias. Our analysis investigates actual voting behavior following actual ticketing events, not reported voting behavior or reported exposure to police ticketing. The administrative data therefore allow us both to sidestep reporting error and to observe the behavior of a quarter-million individuals ticketed over a 6 year period---a far larger pool than even the most robust surveys.

We use individual-level traffic stop data from Hillsborough County, Florida, to identify the turnout patterns of voters who were ticketed between the 2012 and 2018 elections. By matching individual voters who were ticketed to similar voters who were ticketed at later points and running a difference-in-differences model, we estimate the causal effect of these tickets on turnout. We find that being ticketed reduces the chance that an individual will turn out in the subsequent election, but that this effect is smaller for Black voters.

To test the generalizability of our findings in Hillsborough County, we leverage two different datasets that are national in scope. First, we join fees and fines data from municipalities around the country in 2012 and 2017 with a novel dataset estimating racial turnout for thousands of municipalities around the country. While we find that increased fees and fines likely cause lower turnout among non-Black Americans, we find no effect of fees and fines on Black turnout.

Second, we conduct a novel analysis of 2020 national survey data regarding voters’ experience with police stops. We find that Black respondents who reported personal or proximal experience with police stops in the last 12 months are more likely to vote than Black people who did not, while we find no statistically significant relationship between stops and voting among non-Black respondents. While both the municipal-level and survey data present some drawbacks relative to the individual-level administrative data, their national scope provides helpful corroboration for the unique turnout effects of tickets for Black Americans observed in Hillsborough County.

We demonstrate that police ticketing practices---the most widespread form of police contact in America---substantially reduce the turnout of non-Black American voters, but can mobilize Black voters. Our findings complicate existing theories of how criminalization politically socializes Americans, and Black Americans in particular [@Weaver2010]. Additionally, while all forms of criminalization likely contribute to a subjective experience of alienation or group-level exclusion among Black Americans that is well-documented in sociological scholarship [@Bell2017; @Desmond2016], our contribution emphasizes the need for further research regarding how different forms of criminalization affect group-level perceptions of government and resultant political behaviors. Accordingly, we propose a new theory explaining how ticketing may uniquely catalyze Black political participation. Our findings are relevant for interdisciplinary scholars of crime, race, politics, political economy, social movements, and policing.

# Literature Review {-}

An emerging political science literature directly investigates the relationship between criminalization and political behavior. Additionally, sociologists have incorporated contemporary fines and fees practices into a broader empirical-historical theory of racialized punishment policy [@Harris2010; @Friedman2020].

In order to theorize how fines and fees affect political behavior, we consider how criminalization directly shapes the electorate; the existing scholarly understanding of the influence of criminalization on political socialization; and how contemporary ticketing practices are tied to a longer legacy of racist exploitation and thus ripe for political mobilization around narratives of racial injustice.

## Criminalization and political socialization {-}

The American criminal legal system directly shapes the American electorate and excludes voters. Today, more than five million Americans are legally barred from voting due to a felony conviction, and one in every sixteen Black Americans of voting age is disenfranchised [@Uggen2020]. Notably, fines and fees legally disenfranchise Americans as well: 48 states and the District of Columbia authorize some form of wealth-based penal disenfranchisement [@Colgan2019]. Many states require payment of fees as a condition of criminal legal supervision or the payment of all legal financial obligations as a condition of completing a sentence, and failure to comply with supervision or sentence completion conditions can be a barrier to voting rights restoration.

Beyond de jure disenfranchisement, the spatial and racial concentration of criminalization influences affected communities’ perceptions of government. Monica Bell calls this process “legal estrangement,” a conceptual framework meant to capture those perceptions and cultural attitudes (“legal cynicism”) as well as the historical-structural conditions that produced them [@Bell2017]. Research on legal cynicism has found that public perceptions of abusive police practices can reduce willingness to report crimes or cooperate with law enforcement [@Desmond2016; @Tyler2014]. In short, the “hidden curriculum” [@Justice2014; @Meares2017] of the criminal legal system teaches Americans about their identities as citizens---even parts of their identities that have little to do with policing or incarceration. New evidence related to ticketing and legal estrangement has recently emerged: in a survey of residents in three Georgia cities that rely heavily on fines and fees revenue, residents who were ticketed reported lower levels of trust in police and government compared to residents who were not ticketed [@CarpenterII2019].

Because police are the most visible agents of the state, perceptions of unfairness in policing can translate to a recognition of actual group-level structural exclusion from public institutions [@Bell2017]. Soss and Weaver explain how criminal legal contact shapes perceptions of government and political participation: “interviewees have looked, not to City Hall, Congress, or political parties, but rather to their direct experiences with police, jails and prisons, welfare offices, courts, and reentry agencies as they sought to ground their explanations of how government works, what political life is like for them, and how they understand their own political identities” [@Soss2017: 574].

Sarah Brayne found related results in her analysis of survey data. As the intensity of criminal legal contact escalates (i.e. stopped, arrested, convicted), people subjected to criminalization are increasingly likely to avoid institutions that they perceive as surveilling, such as financial or educational institutions [@Brayne2014]. She calls this phenomenon “system avoidance,” and while her analysis did not specifically assess voting behavior, her findings demonstrate how criminal legal contact can discourage Americans’ participation in public institutions as well as how the effect may vary alongside the intensity of criminal legal intervention. Similarly, @Remster2018a finds that men who return home from prison are more likely to avoid contact with their childrens’ schools. Vesla Weaver and colleagues helpfully describe this form of self-preserving withdrawal from public institutions as a “strategic retreat” [@Weaver2020].

In addition to influencing perceptions of government and participation in public institutions, recent scholarship indicates that criminalization influences political behavior. Using existing national survey data and extended interviews, Vesla Weaver and Amy Lerman found that trust in government and willingness to vote decrease as individuals progress through increasingly intense levels of criminal legal contact (questioned by police, arrested, convicted, incarcerated) [@Weaver2010]. Weaver and Lerman also investigate the potential role of criminal legal contact in facilitating a collective political consciousness among Black Americans, or the concept of a “linked fate” where “one’s individual opportunities are linked to the entire racial group” [@Lerman2014: 197]. Their interview data lead them to conclude that while criminal legal contact does increase perceptions of linked fate among Black respondents, “a custodial linked fate is not the building block of greater activism and a mobilized racial identity; instead, it is demobilizing.” Interview respondents who had been incarcerated reported that withdrawal from public institutions and democratic participation constituted a form of self-preservation [@Lerman2014].

Lerman and Weaver’s findings have been corroborated using administrative data: Ariel White found that short jail sentences in Harris County, Texas reduced Black defendant turnout in subsequent elections by 13 percentage points---but the effect on white turnout was not statistically significant [@White2019]. Traci Burch found that North Carolina block groups with higher concentrations of people who had experienced incarceration or criminal legal supervision saw substantial reductions in voter turnout, and confirmed her findings at the individual level [@Burch2014]. These effects can also extend to family members, and the political consequences can be quite durable [@Morris2021].

Although most work documents how contact with the criminal legal system causes strategic retreat, recent work from Hannah Walker [-@Walker2020; -@Walker2020a] complicates this narrative. She argues that when Black Americans understand their contact with the criminal legal system in the context of a narrative of racial injustice, such contact can actually mobilize. She explains: “experiences with punitive policies are subject to interpretation and, when understood through narratives of injustice, lead to wholly different participation outcomes than we usually predict” [@Walker2020: 132]. While she finds that this sense of injustice is especially likely to increase political participation in non-voting ways (such as attending a protest or signing a petition), the logic could extend to electoral participation as well given a relevant electoral context, strong injustice narrative, or relatively weak contact with the criminal legal system. Findings presented in @Laniyonu2019 provide support for this concept: concentrated Stop, Question, and Frisk (SQF) policing in New York City neighborhoods was associated with reduced turnout in federal midterm elections but increased turnout in presidential and mayoral primary elections. Laniyonu also uncovered an association regarding candidate choice—he demonstrates higher turnout for a candidate who opposed SQF policing and lower turnout for a candidate who supported it. In other words, low-level police contact mobilized voters to participate in a local election (which could plausibly affect police practices).

Despite the growing body of high-quality scholarship on the democratic implications of the American criminal legal system, we still know remarkably little about the effects of the most routine sort of contact---police stops and tickets that do not result in an arrest.

## Race and fines and fees practices {-}

Fines and fees are situated in a distinct historical legacy of racist exploitation that may uniquely facilitate a collective “sense of injustice” among Black Americans---the sort of collective perception that @Walker2020a argues could prove to be politically mobilizing.

Cedric Robinson’s articulation of racial capitalism offers a useful theoretical framework for interpreting the trajectory of American anti-Black political economy [@Robinson2000]. Brittany Friedman theorizes fines and fees as a “progression of racial capitalism penology,” emphasizing the centrality of carceral immobility and financial capture in contemporary fines and fees practices [@Friedman2020: 177]. Robinson argues that European industrial capitalism continued a longer political-economic trend of differentiating racial groups in order to dispossess them and exploit their labor.

Following the Civil War and the rise of global industrial capitalism, slave patrols were replaced by more formal police forces [@EqualJusticeInitiative2017; @Hinton2021]. The implementation of the Black Codes, convict leasing, and sharecropping systems marked a transition to a new stage of racial capitalism, wherein criminalization provided an easily-exploited labor supply that offset costs that the state incurred by coercing their labor [@Lichtenstein1996; @Mancini1978]. Contemporary fines and fees practices have been directly compared to sharecropping because both structures share “a long-term pattern of financial exploitation based on race… that limits the life-choices of the affected population” [@Blessett2016: 113].

In the 20th century, residential segregation simultaneously fueled the racial wealth gap and ensured that criminalization was both spatially and racially concentrated [@Rothstein2017; @Hinton2021]. At the level of federal policy, racial threat and whites’ attribution of criminality to Blackness [@Muhammad2010] prompted a shift from the War on Poverty to the War on Crime wherein Black communities’ demands for material security and safety were selectively amplified to promote increased and harsher policing and incarceration [@Hinton2017; @Vargas2017].

The 2008 financial crisis solidified the connection between fiscal austerity and policing, prompting a decline in tax revenues that devastated state and local governments---but law enforcement agencies were largely spared from cuts. At the local level, this means that police budgets today constitute about thirty percent of city budgets nationwide, and cities where residents have lower median incomes dedicate a larger proportion of spending to policing [@Friedman2020]. Similarly, municipalities where a larger share of the population is Black raise more fines and fees revenue per-capita than less-Black municipalities [@Sances2017; @UnitedStatesCommissiononCivilRights2017]. Although major urban centers collect high dollar amounts of fines and fees, some of the most egregious ticketing practices measured on a per-capita basis are concentrated in suburbs or smaller cities with municipal courts---Ferguson being a prominent example [@UnitedStatesDepartmentofJusticeCivilRightsDivision2015; @Mughan2020]. In municipalities with substantial Black populations, then, people who are most heavily affected by criminalization are expected to foot the bill for their own mistreatment. @Sances2017, however, finds that the discriminatory imposition of tickets can be mitigated by descriptive representation (i.e. Black elected officials): The positive association between cities’ reliance on fines and fees and the proportion of Black residents is significantly diminished when Black communities are represented by at least one Black legislator [@Sances2017]. This result aligns with recent research demonstrating that states with more Black legislators spend less on corrections budgets; that racial disparities in arrests are reduced in municipalities represented by a majority-nonwhite city council; and that descriptive representation can reduce racial disparities in the criminal legal system  [@Butler2011; @Gunderson2020; @Eckhouse2016]. It thus seems that the ballot box can be a site from which Black Americans can secure some protections against the most egregious forms of revenue extraction.

Racial capitalist political economy imposes fiscal constraints regardless of city officials’ identities or intentions. In their study of fines and fees reliance in Chicago suburbs, Josh Pacewicz and John Robinson, III show how the racialization of municipal opportunity structures prevents majority-Black suburban governments from acquiring sufficient revenue from “good” sources such as sales taxes or property taxes, catalyzing a reliance on fines and fees to fund government services [@Pacewicz2020]. Pacewicz and Robinson III note that racialized municipal opportunity structures are tied to the political-historical processes of neoliberalism and fiscal austerity, an observation that aligns with recent scholarship detailing the origins of fees paid by incarcerated people [@Kirk2020].

Today, fines and fees primarily target Black Americans: traffic stops and searches disproportionately affect Black motorists nationwide [@Pierson2020], and a host of local analyses confirm the national finding [@Dunn2009; @Goncalves2020; @Mughan2020]. In criminal court proceedings, Black defendants tend to face more severe monetary sanctions [@Edwards2020; @Harris2011], and a 2020 Federal Reserve survey found that five percent of white respondents reported “unpaid legal expenses, fines, or court costs,” whereas twelve percent of Black respondents reported such debt [@BoardofGovernorsoftheFederalReserveSystem2020].

Research regarding police stops and searches in response to local fiscal crises shows how decreasing local revenues affect racist police practices. One study of all U.S. city governments found that local fiscal stress was related to increased “revenue-generating arrests” (e.g. arrests involving civil asset forfeiture) and that Black and Hispanic people were arrested at higher rates [@Makowsky2009]. A study of four million North Carolina traffic stops found that fiscal stress caused a decrease in the probability of white drivers being searched, but no decrease for Black drivers [@Shoub2020]. This suggests that higher reliance on fines and fees results in government officials and police targeting individuals who hold less political power---whether those individuals are out-of-state motorists or members of local marginalized communities [@Makowsky2009].

# Theory {-}

As the preceding sections make clear, past work on the political ramifications of criminal legal contact has largely determined that such contact is demobilizing. Scholars have argued that all contact with the criminal legal system causes political withdrawal, and that the severity of the withdrawal varies according to the intensity of the contact. This finding may not hold true for lower-level criminal legal contact, however. We argue that police ticketing may be an exception to this rule for two reasons: (1) police stops are far less disruptive to Americans’ lives than arrests or periods of incarceration, and (2), the injustice narrative framework [@Walker2020a] indicates that the explicit politicization of police stops by the Black Lives Matter movement in recent years should be mobilizing for Black Americans.

Receiving a ticket is a substantively different experience than being arrested or incarcerated. People who have been convicted of a crime and incarcerated are relegated to an alternate political membership status that Reuben Jonathan Miller and Forrest Stuart call “carceral citizenship,” subjecting them to forms of supervision and sanction “that are otherwise unavailable to [people] who have not been accused of a crime” [@Miller2017: 536]. The politically discouraging effects of criminalization are even stronger for people who are convicted of felonies, who lose their right to vote in all but two U.S. states. This “civil death” [@Ewald2002; @Miller2012] explicitly removes Americans from the electorate; the felony conviction becomes “a durable constraint and marker of their citizenship” [@Lerman2014: 133]. The experience of formal legal exclusion from voting forms a powerful piece of the “hidden curriculum” [@Meares2017] that teaches convicted individuals about their identities as citizens.  Finally, the financial and social costs following a period of incarceration [@Bushway2007] can make it less likely that citizens invest the time and energy into registering to vote, learning where their polling place is, and determining which candidates they support.

>>>>

---
title: "Ticketing and Turnout: The Participatory Consequences of Low-Level Police Contact"

# author:
# - Jonathan Ben-Menachem^[PhD Student, Columbia University, Department of Sociology (jb4487@columbia.edu)]
# - Kevin Morris^[PhD Student, CUNY Graduate Center, Department of Sociology (kmorris@gradcenter.cuny.edu)]
# thanks: Earlier versions of this work were presented at the 2021 American Sociological Association Annual Meeting, the 2021 American Political Science Association Annual Meeting, and the Justice Lab at Columbia University. We would like to thank Bruce Western, Joshua Whitford,  Gerard Torrats-Espinosa, Brittany Friedman, Brendan McQuade, Flavien Ganter, Hannah Pullen-Blasnik, Tim Ittner, LeShae Henderson, and Van Tran for their engagement with our paper in its earliest stages. Special thanks to Sam Donahue for identifying the Tampa dataset that became the core of our analysis and eyeballing many, many iterations of analysis results. And finally, we would like to thank Hannah Walker for deep, charitable feedback that pushed us not to undersell our own contribution
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    keep_tex: true
bibliography: "policing_turnout.bib"
link-citations: yes
fontsize: 12pt
header-includes:
    - \usepackage{rotating}
    - \usepackage{setspace}
abstract: |
  The American criminal legal system is an important site of political socialization: scholars have shown that criminal legal contact reduces turnout, and that criminalization pushes people away from public institutions more broadly. Despite this burgeoning literature, few analyses directly investigate the causal effect of *lower-level* police contact on voter turnout. We use two approaches to do so: we leverage both individual-level administrative ticketing data from Hillsborough County, Florida, and responses to the American National Election Studies. We estimate that a traffic stop decreases an individual’s likelihood of voting by 1.7 percentage points, but depressive effects are smaller for Black residents. The survey data, however, show that Black people who report that they or a family member were recently stopped by police are *more* likely to vote than those who reported no such contact. Taken together, our results indicate that low-level contact with police may have participatory consequences.

  |
  |
  **Keywords**: voting, policing, race, punishment, political participation
---

@@

Although scholars have examined the collateral consequences of this increased reliance on fines and fees [@Sances2017; @Pacewicz2020], comparatively few have explored the moment during which such revenue-raising actually occurs: namely, in the individual interactions between residents and the police via the issuance of a ticket. This “moment” of low-level contact has also been relatively understudied by scholars investigating the participatory consequences of contact with the criminal legal system. Work exploring how criminalization directly and indirectly influences political participation has exploded in recent years. Scholars have found that criminal legal contact (i.e. arrest, conviction, incarceration) consistently discourages voting [@Weaver2010; @Burch2011; @White2019a]. Such work has largely focused on the effects of highly disruptive contact with the criminal legal system such as incarceration and felony convictions [@Burch2014; @Lee2014]. While ticketing involves potentially negative interactions with the state, it does not necessarily carry the disruptive consequences of a felony conviction and might thus politicize Americans in unique ways. Building on recent theoretical work from @Walker2020a, this paper theorizes how local police practices affect voting behavior among ticketed individuals and provides precisely estimated evidence of a causal effect.

Our project represents the first use of individual-level administrative data to identify the causal effect of traffic stops on voter behavior. The use of administrative data marks an important step forward in our understanding of how low-level contact with the criminal legal system structures political participation. Past work looking at the individual-level effects of low-level contact has relied on survey or qualitative data [e.g. @Walker2014; @Weaver2010]. Existing research allows for the testing of specific psychological mechanisms and personal interpretations of criminal legal contact, but does not allow us to generalize more broadly. As Weaver and Lerman [-@Weaver2010: 821] note, it may also introduce measurement bias. Our analysis investigates actual voting behavior following actual ticketing events, not reported voting behavior or reported exposure to a traffic stop. The administrative data therefore allow us both to sidestep reporting error and to observe the behavior of a quarter-million individuals stopped over a 6 year period---a far larger pool than even the most robust surveys.

We use individual-level traffic stop data from Hillsborough County, Florida, to identify the turnout patterns of voters who were stopped between the 2012 and 2018 elections. By matching individual voters who were stopped to similar voters who were stopped at later points and running a difference-in-differences model, we estimate the causal effect of these stops on turnout. This borrows from the logic of regression discontinuities in time: conditional on observable characteristics *and* unobservable factors associated with being ticketed, the timing of the ticket on either side of election day is essentially as-if random. We find that being stopped reduces the chance that an individual will turn out in the subsequent election, but that this effect is smaller for Black voters.

To test the generalizability of our findings in Hillsborough County, we conduct a novel analysis of 2020 national survey data regarding voters’ experience with police stops. We find that Black respondents who reported personal or proximal experience with police stops in the last 12 months are more likely to vote than Black people who did not, while we find no statistically significant relationship between stops and voting among non-Black respondents. While the survey data present some drawbacks relative to the individual-level administrative data, their national scope provides helpful corroboration for the unique turnout effects of traffic stops for Black Americans observed in Hillsborough County.

We demonstrate that police ticketing practices---the most widespread form of police contact in America---substantially reduce the turnout of non-Black American voters, but have a unique effect on Black voter turnout. Our findings complicate existing theories of how criminalization politically socializes Americans, and Black Americans in particular [@Weaver2010]. Additionally, while many forms of criminalization have been found to contribute to a well-documented subjective experience of alienation or group-level exclusion among Black Americans [@Bell2017; @Stuart2016; @Desmond2016; @Zoorob2020; @Desmond2020; @Ang2021], our contribution emphasizes the need for further research regarding how different forms of criminalization affect group-level perceptions of government and resultant political behaviors. Our findings are relevant for interdisciplinary scholars of crime, race, politics, municipal finance, and policing.

# Criminal Legal System and Citizen Identity {-}

In recent years, scholars have paid more attention to the political implications of the American criminal legal system. Evolving in part out of a literature focused on street-level bureaucrats, researchers across the social sciences are asking how daily interactions with agents of the carceral state shape Americans’ relationships to themselves, their communities, and their state. As @Soss2017 argues, this “second face” of the state---“the activities of governing institutions and officials that exercise social control and encompass various modes of coercion, containment, repression, surveillance, regulation, predation, discipline, and violence” (567)---is the primary way through which many class-race subjugated communities develop their political knowledge. “[I]nterviewees have looked, not to City Hall, Congress, or political parties,” they write, “but rather to their direct experiences with police, jails and prisons, welfare offices, courts, and reentry agencies as they sought to ground their explanations of how government works, what political life is like for them, and how they understand their own political identities” [@Soss2017, 574].

Recent work has explored the production of this political knowledge in increasingly sophisticated ways, using both qualitative and quantitative approaches. @Lerman2014’s pathbreaking book *Arresting Citizenship* set the table for future research. Their project combined extensive interviews of individuals living in areas closely monitored by the criminal legal system with novel survey data to probe the political consequences of what they term the “carceral state.” Their work indicates that citizens nearly uniformly react negatively to police and carceral supervision---from low-level interactions to lengthy periods of incarceration, Americans subjected to the second face of the state apparently withdraw from political participation. They find that trust in government and willingness to vote decrease as individuals progress through increasingly intense levels of criminal legal contact (questioned by police, arrested, convicted, incarcerated) [@Weaver2010]. New evidence related to ticketing and legal estrangement has recently emerged: in a survey of residents in three Georgia cities that rely heavily on fines and fees revenue, residents who were ticketed reported lower levels of trust in police and government compared to residents who were not ticketed [@CarpenterII2019].

Sarah Brayne [-@Brayne2014] found related results in her analysis of survey data. As the intensity of criminal legal contact escalates , people subjected to criminalization are increasingly likely to avoid institutions that they perceive as surveilling, such as financial or educational institutions. She calls this phenomenon “system avoidance,” and while her analysis did not specifically assess voting behavior, her findings demonstrate how criminal legal contact can discourage Americans’ participation in public institutions as well as how the effect may vary alongside the intensity of criminal legal intervention. Similarly, @Remster2018a finds that men who return home from prison are more likely to avoid contact with their childrens’ schools. Vesla Weaver and colleagues describe this form of self-preserving withdrawal from public institutions as a “strategic retreat” [@Weaver2020].

These findings can be situated in a process that sociologist Monica Bell calls “legal estrangement,” which captures those perceptions and cultural attitudes (“legal cynicism”) as well as the historical-structural conditions that produced them [@Bell2017]. Research on legal cynicism has found that public perceptions of abusive police practices can reduce willingness to report crimes or cooperate with law enforcement [@Tyler2014]. The “hidden curriculum” [@Justice2014; @Meares2017] of the criminal legal system teaches Americans about their identities as citizens---even parts of their identities that have little to do with policing or incarceration.

# Administrative Data and the Causal Effects of Criminalization {-}

Many scholars have turned to administrative data in recent years to explore the actual effects of contact with the criminal legal system on political participation in the voting booth. To be sure, administrative data have limitations: they are not ideally suited to probe the psychological, social, relational, and other mechanisms leading some citizens to vote and others to abstain. These limitations, however, are balanced by offering researchers precise estimates of actual---and not just reported---turnout.

Traci Burch [-@Burch2011] was among the earliest scholars to match voter-file records with records detailing citizens’ contact with the criminal legal system. Her work complicated a previously dominant theory postulating that the turnout and party registration of formerly incarcerated individuals was largely driven by their sociodemographic characteristics but not their formal conviction or supervision [see, for instance, @Uggen2002]. She argues that citizens who are convicted of felony offenses are less likely to vote, and less likely to register with a party, than citizens with similar sociodemographic characteristics without such a conviction.

Other scholars have similarly matched corrections records with registered voter files to explore the effect of contact with the second face of the state. @White2019a finds that short periods of incarceration reduce Black, but not white, turnout. @Gerber2017 similarly finds relatively small effects of incarceration on turnout. Meredith and Morse [-@Meredith2014; -@Meredith2015] and @Gerber2015 explore the turnout effect of notifying formerly incarcerated citizens of the legal restoration of their voting rights, with mixed results, while @Morris2021a shows that restoring voting rights to individuals while they are still on parole increases post-supervision turnout. Other work has explored the “spillover” effects of such contact, finding that the political behavior of family members [@White2019] and neighbors [@Burch2014; @Morris2020] can be influenced by indirect contact with incarceration. @Morris2021 argues that these effects are durable, showing that citizens whose household members had been to prison were not disproportionately mobilized by a ballot initiative promising to end felony disenfranchisement.

This burgeoning literature using administrative data has given scholars far greater insight into the participatory consequences of American incarceration. It is, however, virtually silent on the effects of *lower level* contact with the criminal legal system. Far more Americans have low-level contact with the police than will ever spend a night behind bars: just under 20 million Americans experience a traffic stop each year, whereas approximately five million Americans are arrested and jailed each year [@Harrell2020; @Jones2019]. Despite the fact that so many more citizens are subjected to traffic stops, we know comparatively little about its political ramifications, a lack that is particularly glaring given how much we have learned in recent years about higher-level contact.

The one project that has used administrative data to explore the political implications of low-level police contact is @Laniyonu2019, which explores the effects of the Stop, Question, and Frisk (SQF) program in New York City. @Laniyonu2019’s causal estimates rely on a cross-sectional matching model to compare voter turnout in precincts with *high* SQF activity to turnout in precincts with similar sociodemographic characteristics but *low* SQF activity, supplemented with cross-sectional analyses of all precincts in the city. His results paint a complicated picture: his causal estimates find that high levels of SQF increased turnout in the 2008 presidential election, but decreased it in the 2010 midterms. While @Laniyonu2019 marks an important step forward in the investigation of the consequences of policing for political behavior, it faces certain data and methodological limitations. Because the SQF data does not include identifying information about the individuals stopped, he cannot explore the turnout of New Yorkers who were actually subjected to SQF. Thus, as he notes in the conclusion, “the ability of the analysis to make causal claims is limited” (554) by the cross-sectional matching approach.

In this project, we seek to extend our understanding of the political consequences of routine contact with the state’s second face. Combining an individual-level administrative panel dataset with national survey data gives us new insight into how traffic stops can shape citizens’ engagement with their democracy.

# How Police Stops Might Structure Turnout {-}

Much of the existing work on the turnout consequences of the criminal legal system points to a process of political learning or political socialization. @Lerman2014 finds that individuals in highly surveilled neighborhoods learn that the government does not care about them, and is uninterested in soliciting their participation. They note that more than half of individuals stopped by the police report agreeing with the statement “the leaders in government care very little about people like me,” compared with just a third of respondents with no police contact (150). The numbers, however, were far higher for individuals convicted of a crime, of whom three-quarters agreed with the statement. Learning about the criminal legal system takes place over a long period of time. Could isolated police stops that neither require sustained contact with the criminal legal system nor (generally) impose punishments as severe as incarceration politically socialize Americans to the same degree as incarceration?

On one hand, it might seem unlikely that a single stop would produce much new knowledge about the state. For many Americans, a traffic stop amounts to little more than a slap on the wrist. To be sure, fines and fees can wreak havoc on the lives of some drivers. If a ticket leads to a suspended driver’s license, the initial stop can snowball into a much bigger life event that could jeopardize employment, or lead to shorter stints of incarceration for individuals apprehended driving on a suspended license. Searches conducted during traffic stops may also lead to arrest or incarceration if a police officer finds contraband in the vehicle. These comparatively less frequent cases might have consequences more akin to those associated with a brief period of incarceration that can also threaten employment. Nevertheless, the average traffic stop is certainly less disruptive than the average period of incarceration, perhaps producing less political knowledge and therefore impacting turnout less.

On the other hand, a far larger set of Americans are stopped by the police than sent to jail or prison each year. According to the American National Election Study, a nationally-representative survey we use later in this study, the incomes and education levels for individuals stopped by the police are meaningfully higher than for individuals who report ever having been arrested. Motorists stopped by police are apparently systematically different compared to people who experience incarceration. Traffic stops might thus be the primary way some of these Americans learn about the criminal legal system. If these Americans have not already “learned” about the system from their neighborhoods or family members, the political consequences of such newly gleaned knowledge might be large.

Traffic stops might provide early insights about the criminal legal system even for Americans who live in more heavily-policed areas. As @Gerber2017 notes in their study arguing that the participatory consequences of incarceration might be small, incarceration “is an outcome that often follows a long series of interactions with the criminal justice system” (1145). In other words, much of what the criminal legal system “teaches” might have already been learned by the time an individual is sent to prison. Someone who is stopped by the police, however, might have had fewer negative interactions with the state. Put differently, stopped individuals might have more to learn than incarcerated individuals with a long history of police contact, resulting in higher turnout effects.

# Potential for Racially Disparate Effects {-}

The potential for more “learning” from a police stop seems especially salient for Black Americans, particularly in recent years. National news over the 2010s was filled with the names of Black Americans who were killed during routine police stops. The Black Lives Matter movement gained national attention in the aftermath of the Ferguson uprising, an uprising sparked by the police killing of Michael Brown but also fueled by that city’s racial discriminatory ticketing practices. In the fall of 2021, *The New York Times* ran an extensive report exploring police killings that occur during traffic stops, providing more proof of their racially disparate nature [@Levenson2021].

We expect that the participatory consequences of such a stop will be different for Black than non-Black Americans. First, traffic stops differ in meaningful ways for Black and non-Black Americans. As @Baumgartner2018 notes, Black Americans are more likely than whites to receive both “light” (i.e., a warning without a ticked) *and* “severe” (i.e., arrest) outcomes from a traffic stop. Although this may seem paradoxical at first, the authors explain: “while many might rejoice in getting a warning rather than a ticket, the racial differences consistently apparent in the data suggest another interpretation for black drivers: even the officer recognized that there was no infraction” [88]. Black drivers are also more likely to be searched and arrested, but less likely to be found with contraband [@Baumgartner2018]. Similarly, @Epp2014 argues that traffic stops are particularly instructive for Black Americans, as pretextual traffic stops politically socialize Black voters to the specific context of discriminatory police ticketing.

The politicization of a police stop can occur differently for Black Americans through other avenues as well. The BLM movement has raised the salience of racially disparate policing across the country, as have the tragic stories of individuals like Philando Castile who was killed during a police stop. Routine interactions with the police are also more likely to turn deadly for Black Americans than for others [@Levenson2021]. Thus, even if an individual police stop for a Black American is relatively unremarkable on its own, the background context that the interaction *could* have turned deadly is likely to result in some level of trauma for Black drivers. Recent work from Hannah Walker [-@Walker2020; -@Walker2020a] argues that this could lead to a *mobilizing* effect, if stopped drivers understand their contact with the criminal legal system in the context of a narrative of racial injustice, such contact can actually mobilize. She explains: “experiences with punitive policies are subject to interpretation and, when understood through narratives of injustice, lead to wholly different participation outcomes than we usually predict” [@Walker2020, 132]. While she finds that this sense of injustice is especially likely to increase political participation in non-voting ways (such as attending a protest or signing a petition), the logic could extend to electoral participation as well given a relevant electoral context, strong injustice narrative, or relatively weak contact with the criminal legal system.

Given that police stops are more likely to be traumatic for Black motorists (because traffic stops come with a threat of escalation into racialized police violence, because they received harsher punishments, *or* because the stop itself was discriminatory) *and* that there exists a framework available for translating negative experiences with the criminal legal system into political action, we expect police stops to have unique participatory consequences for Black Americans.

At the same time, the average Black American already knows far more about the criminal legal system due to the extreme racial disparities in policing and incarceration. As one 2015 study observed, 44% of Black women (and 32% of Black men) have a family member in prison. These numbers compare with just 12% and 6% for white women and men, respectively [@Lee2015]. Black Americans’ disproportionate contact with the criminal legal system is a firmly established fact at this point. This greater baseline exposure could mean that Black Americans have less to “learn” from a marginal police stop, resulting in smaller turnout effects.
